---
title: "Diamonds Dataset Exploratory Analysis"
author: 
  - name: "İsmail İbican"
    affiliation: "Software Engineering, Muğla Sıtkı Koçman University"
date: today
format:
  html:
    theme: 
      light: flatly
      dark: darkly
    toc: true
    toc-title: "📘 Table of Contents"
    toc-location: left
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-summary: "Show/Hide R Code"
    smooth-scroll: true
    page-layout: full
    highlight-style: atom-one-dark
    mainfont: "Nunito"
    fontsize: 1.05em
    fig-cap-location: bottom
    linkcolor: "#0099cc"
    backgroundcolor: "#f8f9fa"
  pdf:
    toc: true
    number-sections: true
    mainfont: "Roboto"
    fontsize: 11pt
    geometry: 
      - margin=1in
    colorlinks: true
    linkcolor: blue
    highlight-style: tango
execute:
  echo: true
  warning: false
  message: false
editor: visual
comments: false
jupyter: false
---

# Brief Overview: Types, Summaries, Missing Values, and Outliers

------------------------------------------------------------------------

```{r}
# Load required libraries
data("diamonds")
library(ggplot2)
library(dplyr)
library(psych)
set.seed(42)
```

> We load the necessary libraries (`ggplot2` for plotting, `dplyr` for data manipulation, `psych` for detailed summaries).\
>
> Then we load the `diamonds` dataset.

------------------------------------------------------------------------

```{r}
# Check structure and summary
str(diamonds)
summary(diamonds)
describe(diamonds)
colSums(is.na(diamonds))
```

------------------------------------------------------------------------

> We inspect the **structure** of the dataset to see variable types,\
>
> use `summary()` for basic statistics, `describe()` for detailed numeric summaries,\
>
> and `colSums(is.na())` to confirm there are no missing values.

------------------------------------------------------------------------

```{r}
# Outlier detection using IQR
numeric_cols <- sapply(diamonds, is.numeric)
for (col in names(diamonds)[numeric_cols]) {
  Q1 <- quantile(diamonds[[col]], 0.25)
  Q3 <- quantile(diamonds[[col]], 0.75)
  IQR <- Q3 - Q1
  outliers <- sum(diamonds[[col]] < (Q1 - 1.5*IQR) | diamonds[[col]] > (Q3 + 1.5*IQR))
  cat(col, ":", outliers, "potential outliers\n")
}

```

------------------------------------------------------------------------

> We detect potential **outliers** in numeric variables using the **IQR method**.\
>
> This helps identify unusually large or small values in `price`, `x`, `y`, `z`, etc.

------------------------------------------------------------------------

# Categorical Coding Reasoning

------------------------------------------------------------------------

```{r}
levels(diamonds$cut)
levels(diamonds$color)
levels(diamonds$clarity)

```

------------------------------------------------------------------------

> We examine **factor levels**.
>
> -   `cut` → ordinal, representing quality (Fair \< Ideal)
>
> -   `color` → ordinal, D is best
>
> -   `clarity` → ordinal, I1 is worst, IF is best\
>     This explains why these variables are treated as **ordinal factors**

------------------------------------------------------------------------

# Transformations (Log–Log Model)

------------------------------------------------------------------------

```{r}
diamonds <- diamonds %>%
  mutate(log_price = log(price),
         log_carat = log(carat))
cor(diamonds$price, diamonds$carat)
cor(diamonds$log_price, diamonds$log_carat)

```

------------------------------------------------------------------------

> We apply a **log transformation** to `price` and `carat` to linearize their exponential relationship.\
>
> After transformation, the correlation increases, confirming a more linear relationship suitable for modeling.

------------------------------------------------------------------------

# Plots with Titles, Axes, and Captions Scatter Plot: log(Price) vs log(Carat)

------------------------------------------------------------------------

```{r}
ggplot(diamonds, aes(x = log_carat, y = log_price, color = cut)) +
  geom_point(alpha = 0.3, size = 1) +
  labs(
    title = "Relationship Between log(Price) and log(Carat)",
    x = "log(Carat)",
    y = "log(Price)",
    color = "Cut Quality",
    caption = "Log transformations linearize the exponential price–carat relationship."
  ) +
  theme_minimal()

```

------------------------------------------------------------------------

> A scatter plot showing **log-transformed price vs carat**, colored by cut.\
>
> The log transformation **helps visualize a linear trend** for modeling purposes.

------------------------------------------------------------------------

## Boxplot: Price by Clarity

------------------------------------------------------------------------

```{r}
ggplot(diamonds, aes(x = clarity, y = price, fill = clarity)) +
  geom_boxplot(outlier.color = "red", alpha = 0.7) +
  labs(
    title = "Distribution of Diamond Prices by Clarity",
    x = "Clarity Grade",
    y = "Price (USD)",
    caption = "Higher clarity grades generally correspond to higher prices."
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

------------------------------------------------------------------------

> Boxplots show the **spread of price across clarity grades**, highlighting how higher clarity generally increases price.\
>
> Outliers are marked in red.

------------------------------------------------------------------------

## Bar Plot: Count by Color

------------------------------------------------------------------------

```{r}
ggplot(diamonds, aes(x = color, fill = color)) +
  geom_bar() +
  labs(
    title = "Count of Diamonds by Color Grade",
    x = "Color Grade (D = Best)",
    y = "Count of Diamonds",
    caption = "Most diamonds fall within mid-range color grades (G–H–I)."
  ) +
  theme_minimal()

```

> Bar plot showing the **number of diamonds in each color grade**, useful for understanding dataset distribution.

------------------------------------------------------------------------

## Histogram: Price Distribution

------------------------------------------------------------------------

```{r}
ggplot(diamonds, aes(x = price)) +
  geom_histogram(bins = 50, fill = "skyblue", color = "white") +
  labs(
    title = "Histogram of Diamond Prices",
    x = "Price (USD)",
    y = "Frequency",
    caption = "Price distribution is right-skewed; many affordable diamonds, few very expensive ones."
  ) +
  theme_minimal()

```

------------------------------------------------------------------------

> Histogram visualizes the **skewed distribution of prices**, showing that most diamonds are in a lower price range.

# Summary

------------------------------------------------------------------------

> -   Dataset has **no missing values**.
>
> <!-- -->
>
> -   Categorical variables are **ordinal** and represent quality levels.
>
> <!-- -->
>
> -   Log–log transformation improves the **linear relationship between carat and price**.
>
> <!-- -->
>
> -   Plots are annotated for clarity with **titles, axes, and captions**, making the report readable and self-contained.
