---
title: "Basics Of R"
subtitle: "Introduction of R "
author: "Ismail Ibican"
date: "last-modified"
date-format: "DD MMMM YYYY"
description: "This document explains the basic concepts and functions of R programming."

format:
  html:
    lang: en
    embed-resources: true   
    smooth-scroll: true     
    link-external-newwindow: true 
    theme:
      light: zephyr   
      dark: darkly   
    toc: true
    toc-location: left
    toc-title: "Table of Contents"
    toc-depth: 3
    toc-expand: true   
    number-sections: true  
    code-fold: true
    code-summary: "Show/Hide Code"
    code-copy: true
    code-overflow: scroll
    grid:
      body-width: 1000px
      margin-width: 300px
    callout-appearance: minimal 
---

```{r}
"Hello World ! "
```

> There is one way to print something else

```{r}
print("Hello World!")
```

### Setting and getting working directory

```{r}
#getwd()
#setwd()
```

### Basics : The class in a nutshell - Everything we'll do comes down to applying functions to adata

# Operations

```{r}
x <- 7 + 5 # Addition
```

```{r}
7 - 5 # Substraction
```

> It goes like that.

```{r}
exp(1)
```

```{r}
cos(3.14593)
```

```{r}
log2(1)
```

# Types

```{r}
typeof(7)
# Output "double"
```

```{r}
is.numeric(7)

# Output : TRUE
```

```{r}
is.na(7)

# Output : FALSE
```

```{r}
is.character(7)

# Output : FALSE
```

### Variables We can give names to data objects; these give us variables. A few variables are built in

```{r}
pi
```

```{r}
letters
```

> Months in English

```{r}
month.abb
```

```{r}
month.name
```

> To find every variable in the document we use ls() function

```{r}
ls()
```

> Removing variables :

```{r}
rm("x")
ls()
```

> Vectors are just able to hold same type of variables or data
>
> And also it's a first data structre type that we will use in R

```{r}
y <- c(1:10)
y

# Seperating vector with a colon that means take all numbers from 1 to 10 both interval is included to output
```

```{r}
y ^ y
```

```{r}
dim(x) <- c(2,5)
x <- c("Hello", "World","!")
x
```

> dim() function is a built in function that turns parameter vector to a matrix.
>
> In this example dim() function turns x vector to a matrix that have 2 rows and 5 columns

### Usage of rep() and seq() functions

```{r}
Example1 <- seq(2:50)

# That means there will be 2 rows and 50 number
```

```{r}
Example1 <- seq(1,9, by=2)
Example1 

# Take the numbers from 1 to 9 with steps 2
```

```{r}
Example1 <- rep(1:2, c(10,15))

# This rep function says that take 1 to 2 and repeat them in order to vector
```
